A method for defining dispersed community territories

### 2024-03-06 R/community-clustering.qmd
minor edits

gg <- ggplot() +
  geom_hline(
    yintercept = 0,
    color = "gray75",
    linewidth = 0.2
  ) +
  geom_line(
    data = commutes, 
    aes(distance/60, log(observed/expected), color = region), 
    linewidth = 0.9,
    lineend = "round"
  ) +
  geom_point(
    data = q95,
    aes(distance/60, y, color = region),
    size = 2
  ) +
  geom_text(
    data = q95,
    aes(distance/60, y, color = region, label = label),
    vjust = 0.5,
    hjust = 0,
    nudge_x = 2,
    size = 11.5/.pt
  ) +
  scale_color_manual(name = NULL, values = region_colors) +
  annotate(
    "point",
    x = min(q95[["distance"]])/60,
    y = 3.0,
    color = "gray25",
    size = 2
  ) +
  annotate(
    "text",
    label = "Observed 95% Quantile",
    x = min(q95[["distance"]])/60 + 2,
    y = 3.0,
    color = "gray25",
    vjust = 0.5,
    hjust = 0,
    size = 11.5/.pt
  ) +
  labs(
    x = "Commute to Nearest Center [mins]",
    y = "Density\nlog(Observed) - log(Random)"
  ) +
  scale_x_continuous(
    breaks = seq(0, 120, by = 30), 
    expand = expansion(0.03)
  ) +
  coord_cartesian(xlim = c(0, 120)) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_line(linewidth = 0.2, color = "gray85")
  )

>>>>

gg <- ggplot() +
  geom_hline(
    yintercept = 0,
    color = "gray75",
    linewidth = 0.2
  ) +
  geom_line(
    data = commutes, 
    aes(distance/60, log(observed/expected), color = region), 
    linewidth = 0.9,
    lineend = "round"
  ) +
  geom_point(
    data = q95,
    aes(distance/60, y, color = region),
    size = 2
  ) +
  geom_text(
    data = q95,
    aes(distance/60, y, color = region, label = label),
    vjust = 0.5,
    hjust = 0,
    nudge_x = 2,
    size = 11.5/.pt
  ) +
  scale_color_manual(name = NULL, values = region_colors) +
  annotate(
    "point",
    x = min(q95[["distance"]])/60,
    y = 3.0,
    color = "gray25",
    size = 2
  ) +
  annotate(
    "text",
    label = "Observed 95% Quantile",
    x = min(q95[["distance"]])/60 + 2,
    y = 3.0,
    color = "gray25",
    vjust = 0.5,
    hjust = 0,
    size = 11.5/.pt
  ) +
  labs(
    x = "Commute to Nearest Center [mins]",
    y = "Probability Density\nlog(Observed) - log(Random)"
  ) +
  scale_x_continuous(
    breaks = seq(0, 120, by = 30), 
    expand = expansion(0.03)
  ) +
  coord_cartesian(xlim = c(0, 120)) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_line(linewidth = 0.2, color = "gray85")
  )

### 2024-03-06 manuscript/manuscript.qmd
minor edits

---
title: A method for defining dispersed community territories
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Scott G. Ortman
    affiliations:
      - ref: colorado
abstract: |
  The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production. However, investigating trade-offs between them requires estimates of the extent of community territories, including their nearby arable land. Here we demonstrate a simple algorithm to do that. Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of modern metropolitan areas, we rely on community centers (or areas of known and persistent interaction between unrelated individuals). We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande, and Cibola regions. A sensitivity analysis is also conducted to demonstrate how each tuning parameter contributes to the algorithm.
keywords:
  - Aggregation
  - Core-Based Statistical Area
  - Community Detection
  - Complex Systems
  - Least-Cost Path Analysis
  - Population Density
target-journal: Journal of Archaeological Science
bibliography: bibliography.bib
corresponding_author: Kenneth.Vernon@colorado.edu
linenumbers: false
format: 
  simple-pdf: default
  docx: default
execute: 
  echo: false
  error: false
  # warning: false
engine: knitr
---

@@

Urban geographers recognize density as a crucial property of human communities
which influences many other properties, including productivity, resource needs,
and infectious disease rates [@angel_etal2021; @bettencourt2021;
@duranton_puga2020]. The archaeological record of subsistence farming societies
contains physical traces of community formations with a wide range of densities,
often glossed as "dispersed" or "aggregated" [@birch2013; @drennan_etal2015;
@gyucha2019]. The primary difference between the two has to do with the spatial
distribution of residents relative to primary (food) production. In dispersed
formations, households are scattered across an area, interspersed with the land
that they farmed, thus requiring them to commute to central places for various
forms of social production. In aggregated formations, in contrast, households
are clustered in a village or town, such that they reside in locations of social
production, but must commute to fields for primary production.

@@

In this paper, we present a method we developed to solve such problems, which
enables us to compare the properties and resilience of dispersed vs. aggregated
communities on the same landscapes. We draw on three extensive compilations of
archaeological data from different portions of the ancestral Pueblo region of
the US Southwest, using the locations of what we call community centers as
tethers for defining territories that we infer were used by social communities
over extended periods of time [@glowacki_ortman2012; @varien_etal2007]. In some
regions these are sites containing more households than could plausibly be
related by reckoned kinship or some form of civic architecture. The key feature
of these centers is that they represent locations against which travel time
distances to other sites can be compared.

While not strictly required by our algorithm, we make the simplifying assumption
that once established, community territories were relatively fixed in space for
the duration of their associated sites. This allows us to use the distribution
of all recorded residences, regardless of their periods of occupation, in
defining community territories. There are several reasons why we believe this is
a reasonable simplification. First, there are strong cross-cultural regularities
in the time individuals spend in daily travel from and to their residences, so
dispersed farming communities tend to be dispersed across the distance
associated with this typical travel time, regardless of the number of households
involved [@marchetti1994]. Second, the archaeological record of our study areas
supports the idea that community territories did not shrink when their
populations aggregated into villages. In some cases, farmsteads were converted
to field houses, with some building materials being reused elsewhere and
evidence for continued limited activity use [@varien1999; @varien1999a;
@varien2002]. In other cases, grid gardens were constructed directly within
older rubble mounds by residents of newer, aggregated villages nearby
[@gauthier_herhahn2005]. Third, an analysis that allowed community territories
to vary by time step would reduce variation in the density of the resulting
units, thus washing out variation in one of the most important social properties
one would ideally want to examine. Fourth, in areas where there is strong survey
coverage, dispersed residences form clusters, often but not always centered on a
few larger settlements, and this clustering is apparent even when the dispersed
residences are plotted by time step [@schachner2012; @varien2002]. Fifth, the
effects of population growth for dispersed farming communities are not parallel
to the effects of population expansion in aggregated settlements. In the latter,
the area of the settlement must grow somewhat because the residential density is
already relatively high. But in dispersed settlements, there is often unused
land between existing farmsteads that can be settled. This, combined with the
strong cross-cultural regularity in typical commute times, leads to much greater
consistency in the spatial extent of residents in a dispersed community than in
the case of an aggregated community. Finally, given the inherent incompleteness
of the archaeological record, it seems most appropriate to use all remains to
define what is effectively the maximum extent of community territories. While it
was undoubtedly true that individual community territories expanded or
contracted over time, the assumption of constant area seems more reasonable than
the assumption of constant density.

@@

So far, we have only grouped site points into clusters or communities. The goal,
however, is to define their spatial extent. That means, for each set of site
points, we need some way of drawing a polygon that encompasses all of them. This
can be done in a number of ways. The simplest strategy would be to find the
centroid of the set of points and draw a circle with the smallest radius that
still includes them all. This would be utterly arbitrary, however, and frankly
unrealistic, not representing any sort of meaningful boundary given the
landscape. An alternative would be to use the convex hull of the set of points,
but this tends to exaggerate the total area of the community, especially when
the distribution of points is suggestive of a concave shape, like the letter
'C'. A concave hull would handle peculiar shapes, but it would also introduce
the opposite problem, restricting the area of the community to an unreasonable
degree. Our somewhat brute force way of threading this needle is to identify
paths with shortest commute times between the outer most farms in a community,
extract the vertices from those paths, and then incorporate those into the set
of points used to define the concave hull. The result is a polygon that is
concave as well in the dimension of time, meaning the paths with shortest
commute times between any two farms never leave the community.

@@

The key to our path analysis is to take information about the terrain in our
study regions and convert it first into meaningful estimates of hiking speed and
then into meaningful estimates of travel time. Fortunately, attempts to model
hiking speed across a range of slopes have advanced considerably in recent
years, most notably in @campbell_etal2019. Drawing on a large, crowd-sourced
fitness tracker data set consisting of approximately 420,000 individual hikes,
jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT,
@campbell_etal2019 offer what is easily the most empirically robust estimate of
hiking speed to date. Our application of Campbell's hiking function relies on
coefficients from the fifth percentile of individuals in their sample, using the
Lorentz curve as recommended by Campbell, and assumes that everyone in the areas
of our analysis were largely equal in their walking speed and endurance. This
makes their walking speeds comparable to, but not the same as, what would be
estimated by Tobler's hiking function [@tobler1993].

@@

The mean areas of these communities are 8.68 km$^2$ in CMV, 6.06 km$^2$ in NRG,
and 7.19 km$^2$ in CIB; the mean room counts (for all community centers and
farms) are 647.76 in CMV, 593.49 in NRG, and 766.68 in CIB; and the mean room
densities are 99.69 rooms per km$^2$ in CMV, 219.77 rooms per km$^2$ in NRG, and
238.76 rooms per km$^2$ in CIB. We emphasize that the means in several cases
deviate to a considerable degree from median values (denoted in @tbl-results
with the Greek letter $\eta$), indicating skew in the data and the potential
presence of large outliers. This is illustrated in the boxplots in
@fig-densities, which show the distributions of these and other variables of
interest that can be derived from our analysis and also used to evaluate its
merits (see also @tbl-results).

@@

In addition to showing the community polygons, @fig-communities also offers a
simple demographic profile for each region based on the spatial distribution of
all rooms (their count per square kilometer) across communities irrespective of
time period. For visualization purposes, the log of the density is used, as it
helps to normalize the distribution and, in particular, to rein in large
outliers that swamp the variance. It is important to note, too, that values in
each region are re-scaled to be in the same range, so the figure cannot be used
to estimate absolute differences between regions. Still, the relative
differences within each region can be enlightening, but even they must be
interpreted with caution.

Not surprisingly, virtually all of the Mesa Verde NP communities show high
densities. The community that includes Goodman Point and Sand Canyon Pueblo just
north of Ute Mountain also has a relatively high density, as do several of the
communities west of Yellow Jacket and along the Dolores River. In the NRG, the
highest density community is near the center of the region. It juts out from the
Rio Grande at its confluence with the Pojoaque River and includes the San
Ildefonso and Pojoaque puebloes. Another high density community lies in a
shallow north-south canyon at the base of the Sangre de Cristos. Additional high
density communities are observed on the Pajarito Plateau in Bandelier NM. In the
CIB area, the three densest communities can be found in the far eastern edge of
the study area near El Morro NM, with additional dense communities near Ramah
just to the west of El Morro and Manuelito in the northwest.

@@

Results of the sensitivity analysis are as expected given the role that the main
tuning parameters play in the algorithm (see @fig-sensitivity). Increasing
$D{\text -}join$ makes it easier to join two communities, so you see growth in
the total area of communities and declines in the total number of resulting
communities as it increases. Increasing $P$ makes it harder to join communities,
so you see fewer and smaller communities as it increases. Increasing
$D{\text -}max$ increases the total area of communities and the average commute
time to centers as it serves to filter out distant farmsteads. Importantly,
finer-grained grid resolutions had no meaningful impact on the results of this
analysis, so the level of aggregation we chose is acceptable. One puzzling
result is that the CIB region shows some volatility in its room densities with
respect to the main tuning parameters. The number of CIB communities is also
largely invariant. We comment on this in the discussion.

@@

Because our algorithm requires that we calculate the commute time between each
farm and community center, it also allows us to evaluate potential spatial -
strictly, commute time - interactions between them. @fig-commutes tries to get
at this idea by showing the difference between (i) the log probability density
of commute times from observed farms to their nearest centers and (ii) the log
probability density of commute times from random locations to their nearest
centers, along with the 95% quantiles of the observed distributions. This is
analogous to what @reese_etal2019 refer to as the "null difference cost
distance." One can interpret it as showing where commute times differ from what
one might expect were farm sites distributed at random across the landscape.
Values much larger than zero indicate that there are more farms located at those
commute times from centers than you would expect by chance. Conversely, values
less than zero indicate that there are less farms located at those locations
than one might expect by chance. In either case, these differences tell us that
there is positive covariance between farmsteads and nearby centers at commute
times of less than approximately one hour (the value we assigned to
$D{\text -}max$).

@@

Figure shows for each region the distribution of relevant variables across
communities.
:::

@@

Because the type of community we aim to delineate is a dispersed *farming*
community, we structure our algorithm around the dimensions of food production,
travel time, and social interaction, as these are major constraints on
settlement for maize farmers [@bocinsky_kohler2014; @bocinsky_etal2016;
@vernon_etal2022a; @vernon_etalinpress; @yaworsky_etal2023]. To give just one
example, our algorithm suggests that 50% of CMV communities offer anywhere from
15-50 hectares per farm. For the CIB and NRG regions, that's 25-45 and 15-35
hectares per farm, respectively. This means that they meet the 14 hectare
threshold required for a household to subsist on maize in these regions
[@reese_etal2019; @benson2011; @bocinsky_varien2017]. And that is before taking
into account that the area of the community is held fixed [pace @reese_etal2019]
and all sites are used regardless of time period, so the derived ratios of land
to farms represent minimum possible levels, not the levels actually experienced
by farmers in these communities at any given time.

@@

In this paper, we develop an algorithm that leverages the difference between
farmsteads and community centers to cluster sites into dispersed farming
communities and define their boundaries. This was inspired by the concept of a
CBSA, along with important contributions from @varien1999 and @reese_etal2019.
Using this algorithm will allow us to capture and compare the changing
properties of both dispersed and aggregated communities using consistent
measurements. For instance, we can use uniform probability density analysis of
pottery assemblages from each site [@ortman2016] to apportion rooms to time
slices and thus estimate changes in population density through time. The ability
to compare these properties for communities of varying scales and densities is
crucial for evaluating their relative costs and benefits for human welfare.

>>>>

---
title: A method for defining dispersed community territories
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Scott G. Ortman
    affiliations:
      - ref: colorado
abstract: |
  The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production. However, investigating trade-offs between these different forms of settlement requires estimates of the extent of community territories, including their nearby arable land. Here we demonstrate a simple algorithm to do that. Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of central business districts, we rely on community centers (or areas of known and persistent interaction between unrelated individuals). We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande, and Cibola regions in the US Southwest. A sensitivity analysis is also conducted to demonstrate how each tuning parameter contributes to the algorithm.
keywords:
  - Aggregation
  - Core-Based Statistical Area
  - Community Detection
  - Complex Systems
  - Least-Cost Path Analysis
  - Population Density
target-journal: Journal of Archaeological Science
bibliography: bibliography.bib
corresponding_author: Kenneth.Vernon@colorado.edu
linenumbers: false
format: 
  simple-pdf: default
  docx: default
execute: 
  echo: false
  error: false
  # warning: false
engine: knitr
---

@@

Urban geographers recognize density as a crucial property of human communities
which influences many other properties, including productivity, resource needs,
and infectious disease rates [@angel_etal2021; @bettencourt2021;
@duranton_puga2020]. The archaeological record of subsistence farming societies
contains physical traces of community formations with a wide range of densities,
often glossed as "dispersed" or "aggregated" [@birch2013; @drennan_etal2015;
@gyucha2019]. The primary difference between the two has to do with the spatial
distribution of residents relative to areas of primary (food) production. In
dispersed formations, households are scattered across an area, interspersed with
the land that they farmed, thus requiring them to commute to central places for
various forms of social production. In aggregated formations, in contrast,
households are clustered in a village or town, such that they reside in
locations of social production, but must commute to fields for primary
production.

@@

In this paper, we present a method we developed to solve such problems, which
enables us to compare the properties and resilience of dispersed vs. aggregated
communities on the same landscapes. We draw on three extensive compilations of
archaeological data from different portions of the ancestral Pueblo region of
the US Southwest, using the locations of what we call community centers as
tethers for defining territories that we infer were used by social communities
over extended periods of time [@glowacki_ortman2012; @varien_etal2007]. In some
regions, community centers are sites containing civic architecture, or more
households than could plausibly be related by reckoned kinship. The key feature
of these centers is that they represent locations against which travel time
distances to other sites can be compared.

While not strictly required by our algorithm, we make the simplifying assumption
that once established, community territories were relatively fixed in space for
the duration of the occupations of their associated sites. This allows us to use
the distribution of all recorded residences, regardless of their periods of
occupation, in defining community territories. There are several reasons why we
believe this is a reasonable simplification. First, there are strong
cross-cultural regularities in the time individuals spend in daily travel from
and to their residences, so dispersed farming communities tend to be dispersed
across the distance associated with this typical travel time, regardless of the
number of households involved [@marchetti1994]. Second, the archaeological
record of our study areas supports the idea that community territories did not
shrink when their populations aggregated into villages. In some cases,
farmsteads were converted to field houses, with some building materials being
reused elsewhere and evidence for continued limited activity use [@varien1999;
@varien1999a; @varien2002]. In other cases, grid gardens were constructed
directly within older rubble mounds by residents of newer, aggregated villages
nearby [@gauthier_herhahn2005]. Third, an analysis that allowed community
territories to vary by time step would reduce variation in the density of the
resulting units, thus washing out variation in one of the most important social
properties one would ideally want to examine. Fourth, in areas where there is
strong survey coverage, dispersed residences form clusters, often but not always
centered on a few larger settlements, and this clustering is apparent even when
the dispersed residences are plotted by time step [@schachner2012; @varien2002].
Fifth, the effects of population growth for dispersed farming communities are
not parallel to the effects of population expansion in aggregated settlements.
In the latter, the area of the settlement must grow somewhat because the
residential density is already relatively high. But in dispersed settlements,
there is often unused land between existing farmsteads that can be settled.
This, combined with the strong cross-cultural regularity in typical commute
times, leads to much greater consistency in the spatial extent of residents in a
dispersed community than in the case of an aggregated community. Finally, given
the inherent incompleteness of the archaeological record, it seems most
appropriate to use all remains to define what is effectively the maximum extent
of community territories. While it was undoubtedly true that individual
community territories expanded or contracted over time, the assumption of
constant area seems more reasonable than the assumption of constant density.

@@

So far, we have only grouped site points into clusters or communities. The goal,
however, is to define their spatial extent. That means, for each set of site
points, we need some way of drawing a polygon that encompasses all of them. This
can be done in a number of ways. The simplest strategy would be to find the
centroid of the set of points and draw a circle with the smallest radius that
still includes them all. This would be utterly arbitrary, however, and frankly
unrealistic, not representing any sort of meaningful boundary given the
landscape. An alternative would be to use the convex hull of the set of points,
but this tends to exaggerate the total area of the community, especially when
the distribution of points is suggestive of a concave shape, like the letter
'C'. A concave hull would handle peculiar shapes, but it would also introduce
the opposite problem, restricting the area of the community to an unreasonable
degree. Our somewhat brute force way of balancing these factors is to identify
paths with shortest commute times between the outer most farms in a community,
extract the vertices from those paths, and then incorporate those into the set
of points used to define the concave hull. The result is a polygon that is
concave as well in the dimension of time, meaning the paths with shortest
commute times between any two farms never leave the community.

@@

The key to our path analysis is to take information about the terrain in our
study regions and convert it first into meaningful estimates of hiking speed and
then into meaningful estimates of travel time. Fortunately, attempts to model
hiking speed across a range of slopes have advanced considerably in recent
years, most notably in @campbell_etal2019. Drawing on a large, crowd-sourced
fitness tracker data set consisting of approximately 420,000 individual hikes,
jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT,
@campbell_etal2019 offer what is easily the most empirically robust estimate of
hiking speed to date. Our application of Campbell's hiking function relies on
coefficients from the fifth percentile of individuals in their sample, using the
Cauchy (or Lorentz) distribution as recommended by Campbell, and assumes that
everyone in the areas of our analysis were largely equal in their walking speed
and endurance. This makes their walking speeds comparable to, but not the same
as, what would be estimated by Tobler's hiking function [@tobler1993].

@@

The mean areas of these communities are 8.68 km$^2$ in CMV, 6.06 km$^2$ in NRG,
and 7.19 km$^2$ in CIB; the mean room counts (for all community centers and
farms) are 647.76 in CMV, 593.49 in NRG, and 766.68 in CIB; and the mean room
densities are 99.69 rooms per km$^2$ in CMV, 219.77 rooms per km$^2$ in NRG, and
238.76 rooms per km$^2$ in CIB. We emphasize that the means in several cases
deviate to a considerable degree from median values (denoted in @tbl-results
with the Greek letter $\eta$), indicating skew in the data and the potential
presence of large outliers. This is illustrated in the boxplots in
@fig-densities, which show the distributions of these and other variables of
interest that can be derived from our analysis and also used to evaluate its
merits (see also @tbl-results). Note that these summaries are time-insensitive.
They represent the total archaeological sample in each region.

@@

In addition to showing the community polygons, @fig-communities also offers a
simple demographic profile for each region based on the spatial distribution of
all rooms (their count per square kilometer) across communities irrespective of
time period. For visualization purposes, the log of the density is used, as it
helps to normalize the distribution and, in particular, to rein in large
outliers that swamp the variance. It is important to note, too, that values in
each region are re-scaled to be in the same range, so the figure cannot be used
to estimate absolute differences between regions. Still, the relative
differences within each region can be enlightening, though even they must be
interpreted with caution.

Not surprisingly, virtually all of the Mesa Verde NP communities show high
densities. The community that includes Goodman Point and Sand Canyon Pueblo just
north of Ute Mountain also has a relatively high density, as do several of the
communities west of Yellow Jacket and along the Dolores River. In the NRG, the
highest density community is near the center of the region. It juts out from the
Rio Grande at its confluence with the Pojoaque River and includes the San
Ildefonso and Pojoaque pueblos. Another high density community lies in a shallow
north-south canyon at the base of the Sangre de Cristos. Additional high density
communities are observed on the Pajarito Plateau in Bandelier NM. In the CIB
area, the three densest communities can be found in the far eastern edge of the
study area near El Morro NM, with additional dense communities near Ramah just
to the west of El Morro and Manuelito in the northwest.

@@

Results of the sensitivity analysis are as expected given the role that the main
tuning parameters play in the algorithm (see @fig-sensitivity). Increasing
$D{\text -}join$ makes it easier to join two communities, thus leading to growth
in the total area of communities and declines in the total number of resulting
communities. Increasing $P$ makes it harder to join communities, thus leading to
fewer and smaller communities. Increasing $D{\text -}max$ increases the total
area of communities and the average commute time to centers as it serves to
filter out distant farmsteads. Importantly, finer-grained grid resolutions had
no meaningful impact on the results of this analysis, so the level of
aggregation we chose appears acceptable. One puzzling result is that the CIB
region shows some volatility in its room densities with respect to the main
tuning parameters. The number of CIB communities is also largely invariant. We
comment on this in the discussion.

@@

Because our algorithm requires that we calculate the commute time between each
farm and community center, we can also evaluate potential spatial - strictly,
commute time - interactions between them. @fig-commutes illustrates this idea by
showing the difference between (i) the log probability density of commute times
from observed farms to their nearest centers and (ii) the log probability
density of commute times from random locations to their nearest centers, along
with the 95% quantiles of the observed distributions. This is analogous to what
@reese_etal2019 refer to as the "null difference cost distance." One can
interpret it as showing where commute times differ from what one might expect if
farm sites were distributed randomly across the landscape. Values larger than
zero indicate that there are more farms located at those commute times from
centers than you would expect by chance. Conversely, values less than zero
indicate that there are less farms located at those locations than one might
expect by chance. In either case, these differences tell us that there is
positive covariance between farmsteads and nearby centers at commute times of
less than approximately one hour (the value we assigned to $D{\text -}max$).

@@

Figure shows for each region the distribution of relevant variables across
communities. Note that these are time-insensitive distributions of the samples
in each region.
:::

@@

Because the type of community we aim to delineate is a dispersed *farming*
community, we structure our algorithm around the dimensions of food production,
travel time, and social interaction, as these are major constraints on
settlement for subsistence farmers [@bocinsky_kohler2014; @bocinsky_etal2016;
@vernon_etal2022a; @vernon_etalinpress; @yaworsky_etal2023]. To give just one
example, our algorithm suggests that 50% of CMV communities offer anywhere from
15-50 hectares per farm. For the CIB and NRG regions, the corresponding figures
are 25-45 and 15-35 hectares per farm, respectively. This means that farms in
all three areas meet the 14 hectare threshold required for a household to
subsist on maize in these regions [@reese_etal2019; @benson2011;
@bocinsky_varien2017]. And that is before taking into account that the area of
the community is held fixed [pace @reese_etal2019] and all sites are used
regardless of time period, so the derived ratios of land to farms represent
minimum possible levels, not the levels actually experienced by farmers in these
communities at any given time.

@@

In this paper, we develop an algorithm that leverages the difference between
farmsteads and community centers to cluster sites into dispersed farming
communities and define their boundaries. This was inspired by the concept of a
CBSA, along with important contributions from @varien1999 and @reese_etal2019.
This algorithm allows us to capture and compare the changing properties of both
dispersed and aggregated communities using consistent measurements. For
instance, we can use uniform probability density analysis of pottery assemblages
from each site [@ortman2016] to apportion rooms to time slices and thus estimate
changes in population density through time. The ability to compare these
properties for communities of varying scales and densities is crucial for
evaluating their relative costs and benefits for human welfare.
